LOINC2HPO_EPIC AND SCC

### 2022-03-18 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
update analysis

# lump together sex that can be treated as Unknown
patient_birth_year_sex_distribution <- patient_birth_year_sex_distribution %>% 
  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
  mutate(gender = recode(gender, .missing = 'Unknown')) %>%
  # sample_n(50) %>%
  # ignore ~ 1 million patients whoes birth_year is unknown
  # filter(!is.na(birth_year)) %>%
  # 2 patients have birth_year after 2020, which is clearly error as our database cut off is 2020
  filter(birth_year <= 2020) %>%
  mutate(birth_year_bucket = cut(birth_year, breaks = c(1800, seq(1931, 2021, by = 10)), dig.lab = 4)) %>%
  group_by(birth_year_bucket, gender) %>% 
  summarise(n = sum(n)) %>% 
  arrange(birth_year_bucket)

@@

loinc2hpo_success_metric <- get_loinc2hpo_success_metric(TRUE)

@@

hpo_counts_per_patient <- get_hpo_counts_per_patient(TRUE)
hpo_counts_per_patient <- hpo_counts_per_patient %>% 
  replace_na(list(n_hpo = 0, n_distinct_hpo = 0, n_distinct_abnormal_hpo = 0, n_distinct_abnormal_hpo_infer = 0))
```

@@

tested_phenotype_rank_direct_by_race <- get_tested_phenotype_rank_direct_by_race(force_rerun = FORCE_RERUN)
tested_phenotype_rank_all_by_race <- get_tested_phenotype_rank_all_by_race(force_rerun = FORCE_RERUN)
```

@@

tested_phenotype_rank_direct_by_race %>% 
  select(termid, label, race, n_patient) %>% 
  rename(observed = n_patient) %>% 
  group_by(termid, label) %>% 
  nest() %>% 
  mutate(data = map(data, function(d) d %>% 
                      right_join(countByRace, by = 'race') %>%
                      replace_na(list(termid = first(.$termid), label = first(.$label), observed = 0)))) %>% 
  mutate(chisq = map(data, phenotype_race_chisq_test)) %>% 
  select(-data) %>% 
  unnest(chisq) %>% 
  arrange(-p.value)
```

```{r}
tested_phenotype_rank_all_by_race %>% filter(distancetoroot == 1) %>% 
  select(termid, label, race, n_patient) %>% 
  rename(observed = n_patient) %>% 
  group_by(termid, label) %>% 
  nest() %>% 
  mutate(data = map(data, function(d) d %>% 
                      right_join(countByRace, by = 'race') %>%
                      replace_na(list(termid = first(.$termid), label = first(.$label), observed = 0)))) %>% 
  mutate(chisq = map(data, phenotype_race_chisq_test)) %>% 
  select(-data) %>% 
  unnest(chisq) %>% 
  arrange(-p.value)
```

@@

racial_abnormality_fisher %>% filter(between(lab_year, 2000, 2019)) %>% 
  filter(race == 'Black or African American') %>% 
  filter(termid %in% hpo_level_1_terms$termid) %>% 
  {
    ggplot(., aes(x = lab_year, y = estimate, group = termid, color = termid)) + 
    geom_point() +
    geom_line() +
    scale_color_discrete(breaks = .$termid, labels = str_c(.$termid, .$label, sep = " ")) +
    xlab('Year') +
    ylab('odds ratio') +
    theme_bw() +
    theme(panel.grid = element_blank(),
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10))

>>>>

# lump together sex that can be treated as Unknown
patient_birth_year_sex_distribution <- patient_birth_year_sex_distribution %>% 
  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
  mutate(gender = recode(gender, .missing = 'Unknown')) %>%
  # sample_n(50) %>%
  # ignore ~ 1 million patients whoes birth_year is unknown
  # filter(!is.na(birth_year)) %>%
  # 2 patients have birth_year after 2020, which is clearly error as our database cut off is 2020
  filter(birth_year <= 2020) %>%
  mutate(birth_year_bucket = cut(birth_year, breaks = c(1800, seq(1931, 2021, by = 10)), dig.lab = 4)) %>%
  group_by(birth_year_bucket, gender) %>% 
  summarise(n = sum(n), .groups = 'drop') %>% 
  arrange(birth_year_bucket)

@@

loinc2hpo_success_metric <- get_loinc2hpo_success_metric(force_rerun = FORCE_RERUN)

@@

hpo_counts_per_patient <- get_hpo_counts_per_patient(force_rerun = FORCE_RERUN)
hpo_counts_per_patient <- hpo_counts_per_patient %>% 
  replace_na(list(n_hpo = 0, n_distinct_hpo = 0, n_distinct_abnormal_hpo = 0, n_distinct_abnormal_hpo_infer = 0))
```

@@

get_tested_phenotype_rank_all_by_race_year_sex <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_rank_all_by_race_year_sex.csv')

  if (force_rerun | !file.exists(df_path)){
    df <-  dbGetQuery(dsca, "
              with testable_phenotypes_direct as (
              	select lab.medical_record_number , person.race, person.gender , lab.context_name , lab.lab_date , lab.lab_time , date_part(year, lab_date) as lab_year, lab.loinc , testable.termid , testable.is_tested_directly 
              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
              	join hai_az_prod.person_msdw_2020july person using (medical_record_number)
              --where testable.is_tested_directly 
                ),
                testable_phenotype_direct_freq as (
                	select termid, lab_year, race, gender, count(*) as n_record, count(distinct medical_record_number) as n_patient
                	from testable_phenotypes_direct
                	group by termid, lab_year, race, gender
                )
                select testable.*, hpo.label, hpo.distancetoroot
                from testable_phenotype_direct_freq testable
                join hai_az_prod.hpo hpo on testable.termid = hpo.termid
                order by termid, lab_year, race, gender;")

tested_phenotype_rank_direct_by_race <- get_tested_phenotype_rank_direct_by_race(force_rerun = FORCE_RERUN)
tested_phenotype_rank_all_by_race <- get_tested_phenotype_rank_all_by_race(force_rerun = FORCE_RERUN)
tested_phenotype_rank_all_by_race_year_sex <- get_tested_phenotype_rank_all_by_race_year_sex(force_rerun = FORCE_RERUN)
```

@@

tested_phenotype_rank_direct_by_race %>% 
  select(termid, label, race, n_patient) %>% 
  rename(observed = n_patient) %>% 
  group_by(termid, label) %>% 
  nest() %>% 
  mutate(data = map(data, function(d) d %>% 
                      right_join(countByRace, by = 'race') %>%
                      replace_na(list(termid = first(.$termid), label = first(.$label), observed = 0)))) %>% 
  mutate(chisq = map(data, phenotype_race_chisq_test)) %>% 
  select(-data) %>% 
  unnest(chisq) %>% 
  arrange(-p.value) %>% 
  head()
```

```{r}
tested_phenotype_rank_all_by_race %>% filter(distancetoroot == 1) %>% 
  select(termid, label, race, n_patient) %>% 
  rename(observed = n_patient) %>% 
  group_by(termid, label) %>% 
  nest() %>% 
  mutate(data = map(data, function(d) d %>% 
                      right_join(countByRace, by = 'race') %>%
                      replace_na(list(termid = first(.$termid), label = first(.$label), observed = 0)))) %>% 
  mutate(chisq = map(data, phenotype_race_chisq_test)) %>% 
  select(-data) %>% 
  unnest(chisq) %>% 
  arrange(-p.value) %>% 
  head()
```

@@

p <- racial_abnormality_fisher %>% filter(between(lab_year, 2003, 2019)) %>% 
  #filter(race == 'Black or African American') %>% 
  #filter(race == 'White') %>%
  filter(race %in% c('White', 'Black or African American', 'Asian')) %>% 
  filter(termid %in% hpo_level_1_terms$termid) %>% 
  {
    ggplot(., aes(x = lab_year, y = estimate, group = termid, color = termid)) + 
    geom_point() +
    geom_line() +
    scale_color_discrete(breaks = .$termid, labels = str_c(.$termid, .$label, sep = " ")) +
    #geom_hline(yintercept = c(0.8, 1.2), color = 'red', linetype = 'dotted') +
    annotate(geom='rect', ymin = 0.8, ymax = 1.2, xmin = -Inf, xmax = Inf, fill = 'magenta', alpha = 0.2) +
    geom_hline(yintercept = 1, color = 'black') +
    xlab('Year') +
    ylab('odds ratio') +
    theme_bw() +
    theme(panel.grid = element_blank(), legend.position = 'right',
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10)) +
    facet_wrap(~race, ncol = 1)
  }

ggsave(filename = glue('{CACHE_DIR}/plot_by_phenotype_racial_difference_over_years.pdf'), plot = p, width = 8, height = 8)

p
```

```{r}
RACE_OF_INTEREST <- 'Black or African American'
df <- tested_phenotype_rank_all_by_race_year_sex %>% 
  group_by(termid, label, lab_year, race) %>% 
  summarise(observed = sum(n_patient), .groups = 'drop') %>% 
  right_join(tested_patient_count_by_year_race, by = c('lab_year', 'race')) %>% 
  filter(lab_year == 2019) %>% 
  ##filter(termid == 'HP:0003351')
  filter(termid == 'HP:0000079')

racial_tested_abnormality_fisher <- tested_phenotype_rank_all_by_race_year_sex %>%
  group_by(termid, label, lab_year, race) %>%
  summarise(observed = sum(n_patient), .groups = 'drop') %>%
  right_join(tested_patient_count_by_year_race, by = c('lab_year', 'race')) %>%
  #filter(lab_year == 2019) %>%
  #filter(termid %in% c('HP:0003351', 'HP:0000079')) %>%
  group_by(lab_year, termid, label) %>%
  nest() %>%
  ungroup() %>%
  mutate(fisher_coeff = future_map(data, racial_abnormality_statistical_test)) %>%
  select(-data) %>%
  unnest(fisher_coeff)

racial_tested_abnormality_fisher %>% nrow()

write.csv(racial_tested_abnormality_fisher, file = glue('{CACHE_DIR}/racial_tested_abnormality_fisher.csv'), row.names = FALSE)

p <- racial_tested_abnormality_fisher %>% filter(between(lab_year, 2003, 2019)) %>%
  #filter(race == 'Black or African American') %>%
  #filter(race == 'White') %>%
  filter(race %in% c('White', 'Black or African American', 'Asian')) %>%
  filter(termid %in% hpo_level_1_terms$termid) %>%
  {
    ggplot(., aes(x = lab_year, y = estimate, group = termid, color = termid)) +
    geom_point() +
    geom_line() +
    scale_color_discrete(breaks = .$termid, labels = str_c(.$termid, .$label, sep = " ")) +
    #geom_hline(yintercept = c(0.8, 1.2), color = 'red', linetype = 'dotted') +
    annotate(geom='rect', ymin = 0.8, ymax = 1.2, xmin = -Inf, xmax = Inf, fill = 'magenta', alpha = 0.2) +
    geom_hline(yintercept = 1, color = 'black') +
    xlab('Year') +
    ylab('odds ratio') +
    theme_bw() +
    theme(panel.grid = element_blank(), legend.position = 'right',
        axis.title = element_text(size = 8), axis.text = element_text(size = 8),
        legend.text = element_text(size = 8), legend.title = element_text(size = 10)) +
    facet_wrap(~race, ncol = 1)
  }

ggsave(filename = glue('{CACHE_DIR}/plot_by_tested_phenotype_racial_difference_over_years.pdf'), plot = p, width = 8, height = 8)

p
```

For each race, identify the most over-tested phenotypes between 2003 and 2019. 
```{r}
overly_tested_abnormalities_by_each_race_top10 <- racial_tested_abnormality_fisher %>% 
  ## bring in how many patients were tested for each abnormality
  inner_join(tested_phenotype_rank_all_by_race_year_sex %>% 
              group_by(termid, lab_year, race) %>% 
              summarise(n_record = sum(n_record), n_patient = sum(n_patient), .groups = 'drop'), 
             by = c('termid', 'lab_year', 'race')) %>% 
  filter(race %in% c('White', 'Black or African American', 'Asian')) %>%
  ## focus on data between 2003 and 2019
  filter(between(lab_year, 2003, 2019)) %>% 
  group_by(termid, label, race) %>% 
  summarise(median_estimate = median(estimate), 
            n_record = sum(n_record),
            n_patient = sum(n_patient),
            .groups = 'drop') %>% 
  arrange(desc(median_estimate)) %>%
  ## do not focus on rarely tested abnormalities
  ## here we require at least 17 years X 100 patient/year have been tested for the abnormality
  filter(n_patient > 1700) %>% 
  ## find top 10 within each race
  group_by(race) %>% 
  slice_head(n=10) %>% 
  ## give some nicer names
  rename(median_OR = median_estimate,
         lab_record = n_record,
         patient_count = n_patient)

write.csv(overly_tested_abnormalities_by_each_race_top10, file=glue('{CACHE_DIR}/overly_tested_abnormalities_by_each_race_top10.csv'), row.names = FALSE)

overly_tested_abnormalities_by_each_race_top10 %>% 
  ungroup() %>% 
  #filter(race == 'Asian') %>% 
  rename(OR = median_OR) %>% 
  rename(`Lab Record` = lab_record) %>% 
  rename(`Patient Count` = patient_count) %>% 
  mutate(OR = round(OR, digits = 3)) %>%
  group_by(race) %>% 
  nest() %>% {
    walk2(.$race, .$data, 
          function(race_name, df) sjPlot::tab_df(df, 
                                                 alternate.rows = TRUE, 
                                                 title = glue("Top 10 Overly Tested Lab Phenotype for {race_name}"), 
                                                 use.viewer = FALSE, 
                                                 file = glue('{CACHE_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html')))
  }

race_name = 'Asian'
table_overly_tested_by_race <- function(race_name){
  df <- overly_tested_abnormalities_by_each_race_top10 %>% 
    ungroup() %>% 
    filter(race == race_name) %>% 
    rename(OR = median_OR) %>% 
    rename(`Lab Record` = lab_record) %>% 
    rename(`Patient Count` = patient_count) %>% 
    mutate(OR = round(OR, digits = 3)) %>%
    select(-race)

  sjPlot::tab_df(df, alternate.rows = TRUE, 
                     title = glue("Top 10 Overly Tested Lab Phenotype for {race_name}: 2003 - 2019"), 
                 footnote = 'OR: odds of having received lab tests for the phenotype in the race vs other races; <br>
                phenotypes with less than 1700 tested patients were excluded',
                 show.footnote = TRUE,
                     use.viewer = FALSE, 
                     file = glue('{CACHE_DIR}/table_overly_tested_abnormalities_for_{race_name}_top10.html'))
}
table_overly_tested_by_race(race_name = 'Asian')
table_overly_tested_by_race(race_name = 'Black or African American')
table_overly_tested_by_race(race_name = 'White')
```
